Precise Spiking Motifs in Neurobiological and Neuromorphic~Data

### 2022-12-28 brainsci-2069565_accepted.tex
%MDPI: We removed the italics. Please confirm this revision. These were kept as they correspond to latin locutions

% !TeX TS-program = pdfLaTeX
% !TeX encoding = UTF-8 Unicode
% !TeX spellcheck = en-US
% !BIB TS-program = bibtex
% -*- coding: UTF-8; -*-
% vim: set fenc=utf-8
%
% REV1
%
% https://www.overleaf.com/project/633bf90497b5b325a3ff692e
% https://framateam.org/spikeai/channels/review_polychrony
% https://www.zotero.org/groups/4562620/polychronies
%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\ResponseReviewerTwo}{
We would like to thank reviewer 2 for his very detailed response which were essential in providing this new revision.

@@

Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{riehle_spike_1997}. Interestingly, these authors showed that ``accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{kilavik_long-term_2009} and showed some tuning to movement direction and reaction time~\citep{grammont_spike_2003}. It is important to note that synchronous events tend to lock to spatio-temporal patterns of neural activity called \href{https://en.wikipedia.org/wiki/Local_field_potential}{\hl{LFP}} beta waves~\citep{denker_lfp_2018} and were also extended to larger assemblies using statistical methods~\citep{torre_asset_2016} (see Section~\ref{sec:detection} for further details). \change[Antoine]{Note also that synchronicity may explain some unintuitive results. Indeed, it has been shown that thalamo-cortical synapses are relatively weak compared to the amount of intracortical activity. However, this pathway is sufficient to drive the cortex, as this input is more often synchronously active, allowing it to more efficiently drive cortical neurons}{Synchronicity is also an interesting proposition to explain how the relatively weak thalamo-cortical synapses are able to drive cortical neurons. Among different explanations, including travelling waves}~\citep{ben-yishai_traveling_1997}\add[Antoine]{, a synchronous activity at the synaptic level may be sufficient to elicit activity in the cortex}~\citep{bruno_cortex_2006}.

Theoretically\add[Antoine]{, even if the vertebrate's neural system is not likely to be modeled only by such deterministic connectivity}~\citep{deneve_bayesian_2004, ballard_dual_2011}, it was shown that a simple model may allow the propagation of such synfire chains~\citep{gewaltig_propagation_2001} by considering the dynamics of leaky integrate-and-fire (LIF) neurons in different groups of a similar size. Each neuron of one group is connected by an excitatory synapse to the next. When a pulse is elicited in the first group, this may generate a spike in the next group. Depending on the concentration of synaptic weight values, this new activity may become more or less synchronized with respect to that of the previous pulse (as measured by the standard deviation of spike times within the pulse). Recursively applying this to a sequence of groups within a chain generates either a synfire propagation or not. A simple simulation of synfire propagation is shown in Figure~\ref{fig:diesman}. A crucial aspect of this emergence is explained by the dynamics of the spiking neuron model~\citep{gerstner_time_1995} and, in particular, the balance between excitation and inhibition~\citep{azouz_stimulus-selective_2008}. This balance was, for instance, modeled by feed-forward inhibition, a fine-scaled latency mechanism that is an essential ingredient in modelling so-called push--pull effects in the primary visual cortex~\citep{kremkow_push-pull_2016}. Further models have shown that such synfire chains could be embedded in topographies~\citep{aviel_embedding_2003}, while others used conductance-based neurons with feed-forward inhibition to improve the robustness of the propagation~\citep{kremkow_functional_2010}. In particular, this was implemented as a computational neuroscience benchmark model using the pyNN language~\citep{davison_pynn_2008} both in CPU-based and neuromorphic hardware~\citep{pfeil_six_2013}. % (see Figure~\ref{fig:pynn}).
%
%----------------------------%
%\begin{figure}
%\centering
%\includegraphics[width=.7\textwidth]{figures/fnins-07-00011-g004.jpg}
%\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background noise  is implemented as random Gaussian current. The state space gives %the duration of synfire chains as a function of the protocol's parameters.}\label{fig:pynn}
%\end{figure}
%----------------------------%
%----------------------------%
\begin{figure}[H]
%\centering
\includegraphics[width=\textwidth]{figures/Diesmann_et_al_1999.pdf}
\caption{
  Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{\hl{using Brian}}.
  The model consists of $10$ groups (arranged with the first group represented in the lowest row) of $100$ neurons each. Five pulses with decreasing jitter are generated in the first group around times $10$, $30$, $50$, $70$ and $90$~\ms~(with jitters given by a standard deviation which linearly decreases from $5$ to $1$~\ms). This generates a pulse after a certain processing delay in the second group with a different jitter. While the first two pulses progressively vanish in the following groups, starting from the third input pulse (with a jitter of $3~\ms$), it is propagated to the following groups. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
\end{figure}
%----------------------------%

@@

More recently, novel multi-unit recording techniques have enabled the identification of travelling waves of neural activity in different areas of the cortex~\citep{muller_cortical_2018}. The authors reviewed these findings by considering the mechanisms by which travelling waves are generated, and evaluated their possible roles in cortical function. In particular, spontaneous travelling waves naturally emerge from horizontal fiber time delays and travel through locally asynchronous-irregular states~\citep{davis_spontaneous_2021}. Studies of sensory-evoked neuronal responses often focus on mean spike rates, with fluctuations treated as internally generated noise. However, fluctuations of spontaneous activity, often organized as travelling waves, shape stimulus-evoked responses and perceptual sensitivity. The mechanisms underlying these waves are unknown. Further, it is unclear whether waves are consistent with the low rate and weakly correlated ``asynchronous-irregular'' dynamics observed in cortical recordings. In that paper, the authors describe a large-scale computational model with topographically organized connectivity and conduction delays relevant to biological scales. They find that spontaneous travelling waves are a general property of these networks. The travelling waves that occur in the model are sparse, with only a small fraction of neurons participating in any individual wave. Consequently, they do not induce measurable spike correlations and remain consistent with locally asynchronous irregular states. Further, by modulating the state of the local network, they can shape responses to incoming inputs as observed {\hl{in vivo}}. %MDPI: we remove italic, please confirm. The following highlights are the same.
Such waves also occur in motor areas and~\citet{linden_movement_2022} have recently presented ensemble recordings of neurons in the lumbar spinal cord that indicate that, rather than alternating, the population is performing a low-dimensional ``rotation'' in neural space, in which the neural activity is cycling through all phases continuously during the rhythmic~behavior.

@@

From another viewpoint, there is a substantial literature in neurobiology indicating that brain dynamics often organize into stereotyped sequences, such as synfire chains~\citep{ikegaya_synfire_2004}, packets~\citep{luczak_sequential_2007} or hippocampal sequences~\citep{pastalkova_internally_2008}. %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Explictly decoding spiking motifs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%We have previously seen that neurobiologists have described stereotyped sequences of neural activation, notably in the hippocampus~\citep{pastalkova_internally_2008,villette_internally_2015,malvache_awake_2016}, and that neurons can be activated in the same order across days~\citep{haimerl_internal_2019}. 
Going further, researchers found precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons, {\hl{in vivo} %MDPI: We removed the italics. Please confirm this revision.
} and {\hl{in vitro}}. These patterns repeat after minutes, maintaining millisecond accuracy. Indeed,~\citet{ikegaya_synfire_2004} demonstrated that in cortical activity, one can find a repetition of several motifs in spike activity (duration around 1 s +/$-$ 0.5 s, some events in motifs are of similar size but sometimes absent). These sequences can be specific to a particular layer or column, can be synchronized with network activity oscillation, and can involve several cells. They also demonstrated that these sequences can form super sequences, so-called \emph{\hl{cortical songs}}. It consists of the assembly of several sequences which repeat in a specific order with a compressed timing. {\hl{In vivo}} spontaneous activity also reveals repeating sequences: about 3000 sequences, each involving 3--10 cells out (of about 900), and lasting up to 3 s. Sequences have specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster, and are associated with a structured spatial organization of the neurons that formed them. %In cortical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)
% %----------------------------%
% \begin{figure}
% \centering
% \includegraphics[width=.6\linewidth]{figures/Ikegaya2004zse0150424620001.jpeg}%pcbi.1006283.g001.PNG_L.png}
% \caption{Fig. 1. from~\citep{ikegaya_synfire_2004} repeated motifs of spontaneous synaptic activity \textit{in vitro} and \textit{in vivo}. (A) Repeated motifs of intracellular activity from layer 5 pyramidal neurons in slices. Panels show segments (red) of the same voltage-clamp recording from the same cell, repeating seconds or minutes after the initial occurrence (blue). Arrows indicate timings of repeated PSCs. (i) Upper trace: low--temporal resolution display of spontaneous activity of a neuron. Lower traces: higher resolution display of the repeated motif at indicated regions of the trace (a to c). (ii) Example of a longer motif. (B) Three repetitions of a motif. The top traces show the motifs superimposed on each other (blue, green, and red), the middle traces show these same traces individually, and the bottom traces show temporally magnified regions of the motifs (a to c). (C) Repeated sequences of intracellular current-clamp recordings \textit{in vivo}. Two (i) and three (ii) repetitions of motifs are shown. Shuffle tests were performed on traces (i), a to c, yielding significantly fewer repeats (fig.~S2, P  0.02). In (i), the blue trace is shifted --2.75 mV; in (ii), the blue trace is shifted --1.58 mV, and the green +0.79 mV.}\label{fig:Ikegaya2004}
% \end{figure}
% %----------------------------%

>>>>

% !TeX TS-program = pdfLaTeX
% !TeX encoding = UTF-8 Unicode
% !TeX spellcheck = en-US
% !BIB TS-program = bibtex
% -*- coding: UTF-8; -*-
% vim: set fenc=utf-8
%
% ACCEPTED VERSION
%
% https://github.com/SpikeAI/2022_polychronies-review
% https://framateam.org/spikeai/channels/review_polychrony
% https://www.zotero.org/groups/4562620/polychronies
%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\ResponseReviewerTwo}{
We would like to thank reviewer 2 for his very detailed response which were essential in providing this new revision.

@@

Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{riehle_spike_1997}. Interestingly, these authors showed that ``accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{kilavik_long-term_2009} and showed some tuning to movement direction and reaction time~\citep{grammont_spike_2003}. It is important to note that synchronous events tend to lock to spatio-temporal patterns of neural activity called \href{https://en.wikipedia.org/wiki/Local_field_potential}{LFP} beta waves~\citep{denker_lfp_2018} and were also extended to larger assemblies using statistical methods~\citep{torre_asset_2016} (see Section~\ref{sec:detection} for further details). \change[Antoine]{Note also that synchronicity may explain some unintuitive results. Indeed, it has been shown that thalamo-cortical synapses are relatively weak compared to the amount of intracortical activity. However, this pathway is sufficient to drive the cortex, as this input is more often synchronously active, allowing it to more efficiently drive cortical neurons}{Synchronicity is also an interesting proposition to explain how the relatively weak thalamo-cortical synapses are able to drive cortical neurons. Among different explanations, including travelling waves}~\citep{ben-yishai_traveling_1997}\add[Antoine]{, a synchronous activity at the synaptic level may be sufficient to elicit activity in the cortex}~\citep{bruno_cortex_2006}.

Theoretically\add[Antoine]{, even if the vertebrate's neural system is not likely to be modeled only by such deterministic connectivity}~\citep{deneve_bayesian_2004, ballard_dual_2011}, it was shown that a simple model may allow the propagation of such synfire chains~\citep{gewaltig_propagation_2001} by considering the dynamics of leaky integrate-and-fire (LIF) neurons in different groups of a similar size. Each neuron of one group is connected by an excitatory synapse to the next. When a pulse is elicited in the first group, this may generate a spike in the next group. Depending on the concentration of synaptic weight values, this new activity may become more or less synchronized with respect to that of the previous pulse (as measured by the standard deviation of spike times within the pulse). Recursively applying this to a sequence of groups within a chain generates either a synfire propagation or not. A simple simulation of synfire propagation is shown in Figure~\ref{fig:diesman}. A crucial aspect of this emergence is explained by the dynamics of the spiking neuron model~\citep{gerstner_time_1995} and, in particular, the balance between excitation and inhibition~\citep{azouz_stimulus-selective_2008}. This balance was, for instance, modeled by feed-forward inhibition, a fine-scaled latency mechanism that is an essential ingredient in modelling so-called push--pull effects in the primary visual cortex~\citep{kremkow_push-pull_2016}. Further models have shown that such synfire chains could be embedded in topographies~\citep{aviel_embedding_2003}, while others used conductance-based neurons with feed-forward inhibition to improve the robustness of the propagation~\citep{kremkow_functional_2010}. In particular, this was implemented as a computational neuroscience benchmark model using the pyNN language~\citep{davison_pynn_2008} both in CPU-based and neuromorphic hardware~\citep{pfeil_six_2013}. % (see Figure~\ref{fig:pynn}).
%
%----------------------------%
%\begin{figure}
%\centering
%\includegraphics[width=.7\textwidth]{figures/fnins-07-00011-g004.jpg}
%\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background noise  is implemented as random Gaussian current. The state space gives %the duration of synfire chains as a function of the protocol's parameters.}\label{fig:pynn}
%\end{figure}
%----------------------------%
%----------------------------%
\begin{figure}[H]
%\centering
\includegraphics[width=\textwidth]{figures/Diesmann_et_al_1999.pdf}
\caption{
  Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{using Brian}.
  The model consists of $10$ groups (arranged with the first group represented in the lowest row) of $100$ neurons each. Five pulses with decreasing jitter are generated in the first group around times $10$, $30$, $50$, $70$ and $90$~\ms~(with jitters given by a standard deviation which linearly decreases from $5$ to $1$~\ms). This generates a pulse after a certain processing delay in the second group with a different jitter. While the first two pulses progressively vanish in the following groups, starting from the third input pulse (with a jitter of $3~\ms$), it is propagated to the following groups. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
\end{figure}
%----------------------------%

@@

More recently, novel multi-unit recording techniques have enabled the identification of travelling waves of neural activity in different areas of the cortex~\citep{muller_cortical_2018}. The authors reviewed these findings by considering the mechanisms by which travelling waves are generated, and evaluated their possible roles in cortical function. In particular, spontaneous travelling waves naturally emerge from horizontal fiber time delays and travel through locally asynchronous-irregular states~\citep{davis_spontaneous_2021}. Studies of sensory-evoked neuronal responses often focus on mean spike rates, with fluctuations treated as internally generated noise. However, fluctuations of spontaneous activity, often organized as travelling waves, shape stimulus-evoked responses and perceptual sensitivity. The mechanisms underlying these waves are unknown. Further, it is unclear whether waves are consistent with the low rate and weakly correlated ``asynchronous-irregular'' dynamics observed in cortical recordings. In that paper, the authors describe a large-scale computational model with topographically organized connectivity and conduction delays relevant to biological scales. They find that spontaneous travelling waves are a general property of these networks. The travelling waves that occur in the model are sparse, with only a small fraction of neurons participating in any individual wave. Consequently, they do not induce measurable spike correlations and remain consistent with locally asynchronous irregular states. Further, by modulating the state of the local network, they can shape responses to incoming inputs as observed in biology. Such waves also occur in motor areas and~\citet{linden_movement_2022} have recently presented ensemble recordings of neurons in the lumbar spinal cord that indicate that, rather than alternating, the population is performing a low-dimensional ``rotation'' in neural space, in which the neural activity is cycling through all phases continuously during the rhythmic~behavior.

@@

From another viewpoint, there is a substantial literature in neurobiology indicating that brain dynamics often organize into stereotyped sequences, such as synfire chains~\citep{ikegaya_synfire_2004}, packets~\citep{luczak_sequential_2007} or hippocampal sequences~\citep{pastalkova_internally_2008}. %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Explictly decoding spiking motifs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%We have previously seen that neurobiologists have described stereotyped sequences of neural activation, notably in the hippocampus~\citep{pastalkova_internally_2008,villette_internally_2015,malvache_awake_2016}, and that neurons can be activated in the same order across days~\citep{haimerl_internal_2019}. 
Going further, researchers found precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons, \emph{in vivo} %MDPI: We removed the italics. Please confirm this revision. These were kept as they correspond to latin locutions
 and \emph{in vitro}. These patterns repeat after minutes, maintaining millisecond accuracy. Indeed,~\citet{ikegaya_synfire_2004} demonstrated that in cortical activity, one can find a repetition of several motifs in spike activity (duration around 1 s +/$-$ 0.5 s, some events in motifs are of similar size but sometimes absent). These sequences can be specific to a particular layer or column, can be synchronized with network activity oscillation, and can involve several cells. They also demonstrated that these sequences can form super sequences, so-called \emph{\hl{cortical songs}}. It consists of the assembly of several sequences which repeat in a specific order with a compressed timing. {\hl{In vivo}} spontaneous activity also reveals repeating sequences: about 3000 sequences, each involving 3--10 cells out (of about 900), and lasting up to 3 s. Sequences have specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster, and are associated with a structured spatial organization of the neurons that formed them. %In cortical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)
% %----------------------------%
% \begin{figure}
% \centering
% \includegraphics[width=.6\linewidth]{figures/Ikegaya2004zse0150424620001.jpeg}%pcbi.1006283.g001.PNG_L.png}
% \caption{Fig. 1. from~\citep{ikegaya_synfire_2004} repeated motifs of spontaneous synaptic activity \textit{in vitro} and \textit{in vivo}. (A) Repeated motifs of intracellular activity from layer 5 pyramidal neurons in slices. Panels show segments (red) of the same voltage-clamp recording from the same cell, repeating seconds or minutes after the initial occurrence (blue). Arrows indicate timings of repeated PSCs. (i) Upper trace: low--temporal resolution display of spontaneous activity of a neuron. Lower traces: higher resolution display of the repeated motif at indicated regions of the trace (a to c). (ii) Example of a longer motif. (B) Three repetitions of a motif. The top traces show the motifs superimposed on each other (blue, green, and red), the middle traces show these same traces individually, and the bottom traces show temporally magnified regions of the motifs (a to c). (C) Repeated sequences of intracellular current-clamp recordings \textit{in vivo}. Two (i) and three (ii) repetitions of motifs are shown. Shuffle tests were performed on traces (i), a to c, yielding significantly fewer repeats (fig.~S2, P  0.02). In (i), the blue trace is shifted --2.75 mV; in (ii), the blue trace is shifted --1.58 mV, and the green +0.79 mV.}\label{fig:Ikegaya2004}
% \end{figure}
% %----------------------------%
